---
title: "R Notebook"
output: html_notebook
---

#=======================================================================

# Rattle is Copyright (c) 2006-2018 Togaware Pty Ltd.
# It is free (as in libre) open source software.
# It is licensed under the GNU General Public License,
# Version 2. Rattle comes with ABSOLUTELY NO WARRANTY.
# Rattle was written by Graham Williams with contributions
# from others as acknowledged in 'library(help=rattle)'.
# Visit https://rattle.togaware.com/ for details.

#=======================================================================
# Rattle timestamp: 2018-12-04 22:09:41 x86_64-w64-mingw32 

# Rattle version 5.2.0 user 'Gustavo'

# This log captures interactions with Rattle as an R script. 

# For repeatability, export this activity log to a 
# file, like 'model.R' using the Export button or 
# through the Tools menu. Th script can then serve as a 
# starting point for developing your own scripts. 
# After xporting to a file called 'model.R', for exmample, 
# you can type into a new R Console the command 
# "source('model.R')" and so repeat all actions. Generally, 
# you will want to edit the file to suit your own needs. 
# You can also edit this log in place to record additional 
# information before exporting the script. 
 
# Note that saving/loading projects retains this log.

# We begin most scripts by loading the required packages.
# Here are some initial packages to load and others will be
# identified as we proceed through the script. When writing
# our own scripts we often collect together the library
# commands at the beginning of the script here.

```{r}

library(rattle)   # Access the weather dataset and utilities.
library(magrittr) # Utilise %>% and %<>% pipeline operators.

# This log generally records the process of building a model. 
# However, with very little effort the log can also be used 
# to score a new dataset. The logical variable 'building' 
# is used to toggle between generating transformations, 
# when building a model and using the transformations, 
# when scoring a dataset.

building <- TRUE
scoring  <- ! building

# A pre-defined value is used to reset the random seed 
# so that results are repeatable.

crv$seed <- 42 

#=======================================================================
# Rattle timestamp: 2018-12-04 22:09:50 x86_64-w64-mingw32 

# Load a dataset from file.

fname         <- "file:///C:/Users/Gustavo/Documents/MD2018/orcam_class.csv" 
crs$dataset <- read.csv(fname,
			na.strings=c(".", "NA", "", "?"),
			strip.white=TRUE, encoding="UTF-8")

```

```{r}


#=======================================================================
# Rattle timestamp: 2018-12-04 22:09:50 x86_64-w64-mingw32 

# Action the user selections from the Data tab. 

# Build the train/validate/test datasets.

# nobs=91 train=64 validate=14 test=13

set.seed(crv$seed)

crs$nobs <- nrow(crs$dataset)

crs$train <- sample(crs$nobs, 0.7*crs$nobs)

crs$nobs %>%
  seq_len() %>%
  setdiff(crs$train) %>%
  sample(0.15*crs$nobs) ->
crs$validate

crs$nobs %>%
  seq_len() %>%
  setdiff(crs$train) %>%
  setdiff(crs$validate) ->
crs$test

# The following variable selections have been noted.

crs$input     <- c("X", "Unidade.Or.e7.ament.e1.ria",
                   "Grupo.de.Natureza.e.Despesa", "Empenho2000",
                   "Empenho2001", "Empenho2002", "Empenho2003",
                   "Empenho2004", "Empenho2005", "Empenho2006",
                   "Empenho2007", "Empenho2008", "Empenho2009",
                   "Empenho2010", "Empenho2011", "Empenho2012",
                   "Empenho2013", "Empenho2014", "Empenho2015",
                   "Empenho2016", "Empenho2017",
                   "dota.e7..e3.o.atualizada", "Empenho2018")

crs$numeric   <- c("X", "Empenho2000", "Empenho2001",
                   "Empenho2002", "Empenho2003", "Empenho2004",
                   "Empenho2005", "Empenho2006", "Empenho2007",
                   "Empenho2008", "Empenho2009", "Empenho2010",
                   "Empenho2011", "Empenho2012", "Empenho2013",
                   "Empenho2014", "Empenho2015", "Empenho2016",
                   "Empenho2017", "dota.e7..e3.o.atualizada",
                   "Empenho2018")

crs$categoric <- c("Unidade.Or.e7.ament.e1.ria",
                   "Grupo.de.Natureza.e.Despesa")

crs$target    <- "Classifcadores"
crs$risk      <- NULL
crs$ident     <- NULL
crs$ignore    <- NULL
crs$weights   <- NULL

#=======================================================================
# Rattle timestamp: 2018-12-04 22:10:11 x86_64-w64-mingw32 

# The 'Hmisc' package provides the 'contents' function.

library(Hmisc, quietly=TRUE)

# Obtain a summary of the dataset.

contents(crs$dataset[crs$train, c(crs$input, crs$risk, crs$target)])
summary(crs$dataset[crs$train, c(crs$input, crs$risk, crs$target)])

#=======================================================================
# Rattle timestamp: 2018-12-04 22:10:48 x86_64-w64-mingw32 

# Generate a correlation plot for the variables. 

# The 'corrplot' package provides the 'corrplot' function.

library(corrplot, quietly=TRUE)

# Correlations work for numeric variables only.

crs$cor <- cor(crs$dataset[crs$train, crs$numeric], use="pairwise", method="pearson")

# Order the correlations by their strength.

crs$ord <- order(crs$cor[1,])
crs$cor <- crs$cor[crs$ord, crs$ord]

# Display the actual correlations.

print(crs$cor)

# Graphically display the correlations.

corrplot(crs$cor, mar=c(0,0,1,0))
title(main="Correlation orcam_class.csv using Pearson",
    sub=paste("Rattle", format(Sys.time(), "%Y-%b-%d %H:%M:%S"), Sys.info()["user"]))

#=======================================================================
# Rattle timestamp: 2018-12-04 22:11:36 x86_64-w64-mingw32 

# Build a Random Forest model using the traditional approach.

set.seed(crv$seed)

crs$rf <- randomForest::randomForest(Classifcadores ~ .,
  data=crs$dataset[crs$train, c(crs$input, crs$target)], 
  ntree=1000,
  mtry=4,
  importance=TRUE,
  na.action=randomForest::na.roughfix,
  replace=FALSE)

# Generate textual output of the 'Random Forest' model.

crs$rf

# List the importance of the variables.

rn <- round(randomForest::importance(crs$rf), 2)
rn[order(rn[,3], decreasing=TRUE),]



```

```{r}

#=======================================================================
# Rattle timestamp: 2018-12-04 22:11:36 x86_64-w64-mingw32 

# Build a Random Forest model using the traditional approach.

set.seed(crv$seed)

crs$rf <- randomForest::randomForest(Classifcadores ~ .,
  data=crs$dataset[crs$train, c(crs$input, crs$target)], 
  ntree=1000,
  mtry=4,
  importance=TRUE,
  na.action=randomForest::na.roughfix,
  replace=FALSE)

# Generate textual output of the 'Random Forest' model.

crs$rf

# List the importance of the variables.

rn <- round(randomForest::importance(crs$rf), 2)
rn[order(rn[,3], decreasing=TRUE),]


```

```{r}

# Time taken: 0.80 secs

# Plot the OOB ROC curve.

library(verification)
aucc <- verification::roc.area(as.integer(as.factor(crs$dataset[crs$train, crs$target]))-1,
                 crs$rf$votes[,2])$A
verification::roc.plot(as.integer(as.factor(crs$dataset[crs$train, crs$target]))-1,
         crs$rf$votes[,2], main="")
legend("bottomright", bty="n",
       sprintf("Area Under the Curve (AUC) = %1.3f", aucc))
title(main="OOB ROC Curve Random Forest orcam_class.csv",
    sub=paste("Rattle", format(Sys.time(), "%Y-%b-%d %H:%M:%S"), Sys.info()["user"]))

#=======================================================================
# Rattle timestamp: 2018-12-04 22:12:06 x86_64-w64-mingw32 

# KMeans 

# Reset the random number seed to obtain the same results each time.

set.seed(crv$seed)

```

```{r}
# The 'reshape' package provides the 'rescaler' function.

library(reshape, quietly=TRUE)

# Generate a kmeans cluster of size 10.

crs$kmeans <- kmeans(sapply(na.omit(crs$dataset[crs$train, crs$numeric]), rescaler, "range"), 10)

#=======================================================================
# Rattle timestamp: 2018-12-04 22:12:09 x86_64-w64-mingw32 

# Report on the cluster characteristics. 

# Cluster sizes:

paste(crs$kmeans$size, collapse=' ')

# Data means:

colMeans(sapply(na.omit(crs$dataset[crs$train, crs$numeric]), rescaler, "range"))

# Cluster centers:

crs$kmeans$centers

# Within cluster sum of squares:

crs$kmeans$withinss

```

```{r}

# Time taken: 0.01 secs

#=======================================================================
# Rattle timestamp: 2018-12-04 22:14:04 x86_64-w64-mingw32 

# Reload the project data (variable crs) from file.

load("C:\Users\Gustavo\Documents\MD2018\Atividades\Atividade_13\orcam_class.rattle")

#=======================================================================
# Rattle timestamp: 2018-12-04 22:16:11 x86_64-w64-mingw32 

# Save the project data (variable crs) to file.

save(crs, file="C:/Users\Gustavo\Documents\MD2018\Atividades\Atividade_13\orcam_class.r.rattle", compress=TRUE)

#=======================================================================
# Rattle timestamp: 2018-12-04 22:16:33 x86_64-w64-mingw32 

# Save the project data (variable crs) to file.

save(crs, file="C:/Users\Gustavo\Documents\MD2018\Atividades\Atividade_13\orcam_class.r.Rdata", compress=TRUE)


```
